---
title: "ENVS-193DS_Final"
author: "Iris Carnahan"
date: "2026-06-07"
format: pdf
---
# Link to GitHub repo
https://github.com/iriscarnahan05/ENVS-193DS_spring-2026_final

# Set-up tasks

```{r}
#| label: reading-in-packages

library(tidyverse) # general use
library(here) # cleaning data frames
library(janitor) # file/folder organization
library(readxl) # reading .xlsx files
library(ggeffects) # calculating the data
library(dplyr)
library(forcats)
library(MuMIn)
library(DHARMa)
# reading in `nest_data_final` csv file
nest_data <- read.csv(here("data", "nest_data_final.csv"))
```

# Problem 1

## a. Transparent statistical methods

In part 1, they used a binomial generalized linear model (GLM), or a logistic regression. In part 2, they used a chi-squared test.

## b. Suggestions for rewriting

In the San Francisco Bay, the distance from the water edge tends to predict the likelihood of microhabitat use by American Avocets.

Noticeably, certain shorebirds have shown preference for different habitat types in the San Francisco Bay.

## c. Further analysis

The amount of restored native plant cover (measured in % cover) could influence the probability of American Avocet microhabitat use, because they tend to prefer wide-open areas rather than highly vegetated ones in order to forage and feed. Measuring in percentage makes this variable continuous, and this data could be collected via quadrat estimates.

# Problem 2

## a. Clean and wrangle

```{r}
#| label: cleaning-data
# creating data frame called `nests_clean`
nests_clean <- nest_data |> 
  # selecting columns of interest
  select(height_cm, case_control, ant_species) |> 
  # filtering to only include ant species AB, RRB, and BBR
  filter(ant_species == "AB" |
         ant_species == "RRB" |
         ant_species == "BBR") |> 
  # reordering ant species
  mutate(ant_species = factor(ant_species),
         ant_species = fct_relevel(
         ant_species,
         "AB", "RRB", "BBR")) |> 
  # creating new column to store scientific names
  mutate(scientific_name = case_when(
    ant_species == "AB" ~ "C. sjostedti",
    ant_species == "RRB" ~ "C. mimosae",
    ant_species == "BBR" ~ "C. nigriceps",
    # fills in NA to match with ant species NA
    TRUE ~ NA_character_
  ))
#display 10 rows of object
slice_sample(n = 10,
nests_clean)
```

## b. Response variable

The response variable is the occurrence of a bird nest of any species. The 0 indicates that there is no nest, and the 1 indicates that there is a bird's nest.

## c. Purpose of study

Tree height is a continuous variable measured in centimeters. The research shows in the second paragraph of the results section that the taller the tree, the more likely  it is that a bird will have nested in in it.

Ant species are shown to be a categorical variable, as shown at the end of the first paragraph in the Methods section of the research paper. The third paragraph of the results showed that trees inhabited by *C. mimosae* and *C. nigriceps* were more likely to contain nests.

## d. Table of models

```{r}
#| label: interaction-term

# no interaction
nest_mod1 <- glm(case_control ~ ant_species + height_cm,
             data = nests_clean,
             family = "binomial")
# interaction
nest_mod2 <- glm(case_control ~ ant_species * height_cm,
                data = nests_clean,
                family = "binomial")
```

## e. Run the models

```{r}
#| label: model-fitting

# model 0: null model
model0 <- glm(
  case_control ~ 1, # formula: no predictors
  data = nests_clean) # clean data frame
# model 1: all predictors
model1 <- glm(
  case_control ~ ant_species + height_cm,
  data = nests_clean)
# model 2: ant species only
model2 <- glm(
  case_control ~ ant_species,
  data = nests_clean
)
# model 4: tree height only
model3 <- glm(
  case_control ~ height_cm,
  data = nests_clean
)
```

## f. Select the best model

```{r}
#| label: calculating_aic
# calculating AIC
AICc(
  model0,
  model1,
  model2,
  model3,
  nest_mod2
) |> 
  # arranging models in order of AIC
  arrange(AICc)
```

The best model as determined by Akaike's Information Criterion (AIC) is the one where both predictor variables of ant species and tree height (cm) had an interactive effect on the presence of a nest. Accounting for both predictive variables and the potential effects they have on each other best predicts the data outcome.

## g. Check the diagnostics

```{r}
#| label: check-diagnostics
# displaying plots in 2 by 2 grid
par(mfrow = c(2, 2))
# generating diagnostic plots
plot(nest_mod2)
```

## h. Visualize the model predictions 

```{r}
#| label: create-model-preds
pred <- ggpredict(nest_mod2,
          terms = c("height_cm",
                    "ant_species"))
```


```{r}
#| label: model-preds-visualization

ggplot(data = nests_clean,
       aes(x = height_cm,
           y = case_control)) +
  geom_point(aes(color = ant_species),
             shape = 21) +
  geom_line(data = pred,
            aes(x = x,
                y = predicted,
                color = group),
            linewidth = 2) +
  geom_ribbon(data = pred,
              aes(x = x, 
                  y = predicted,
                  ymin = conf.low,
                  ymax = conf.high,
                  fill = group),
              alpha = 0.1) +
  facet_wrap(~ group,
             labeller = labeller(
               group = c(RRB = "C. mimosae",
                               AB = "C. sjostedti",
                               BBR = "C. nigriceps")
             )) +
  scale_color_manual(values = c("RRB" = "coral",
                                "AB" = "darkolivegreen1",
                                "BBR" = "deeppink2")) +
  scale_x_continuous(breaks = c(100, 200, 300, 400, 500, 600)) + 
  labs(x = "Height (cm)",
       y = "Probability of Birds Nests") +
  theme_classic() +
  theme(legend.position = "none")
```

## i. Write a caption for your figure

**Figure 1. The higher up in the trees with *C. nigriceps* ant populations, the more likely birds will nest.** Each colored line represents the probability of finding a birds nest at a given height (cm). Each colors and graph differentiates the ant population coexisting on the observed trees(green: *C. sjostedti*, orange: *C. mimosae*, pink: *C. nigriceps*), and the faded bands, or ribbons, around the lines represent the 95% confidence interval, or the precision of the predictions. 

## j. Calculate model predictions

```{r}
#| label: applied-model-predictions
# predict occurrence of nests using model 2
ggpredict(nest_mod2,
          terms = c("height_cm [600]", # at 600 cm
                    "ant_species")) # differentiates between ant species
```


## k. Interpret your results

# Problem 3
