---
title: "ENVS-193DS_Final"
author: "Iris Carnahan"
date: "2026-06-07"
format: pdf
---
# Link to GitHub repo
https://github.com/iriscarnahan05/ENVS-193DS_spring-2026_final

# Set-up tasks

```{r}
#| label: reading-in-packages

library(tidyverse) # general use
library(here) # cleaning data frames
library(janitor) # file/folder organization
library(readxl) # reading .xlsx files
library(ggeffects) # calculating the data
library(dplyr)
library(forcats)
# reading in `nest_data_final` csv file
nest_data <- read.csv(here("data", "nest_data_final.csv"))
```

# Problem 1

## a. Transparent statistical methods

In part 1, they used a binomial generalized linear model (GLM), or a logistic regression. In part 2, they used a chi-squared test.

## b. Suggestions for rewriting

In the San Francisco Bay, the distance from the water edge tends to predict the likelihood of microhabitat use by American Avocets.

Noticeably, certain shorebirds have shown preference for different habitat types in the San Francisco Bay.

## c. Further analysis

The amount of restored native plant cover (measured in % cover) could influence the probability of American Avocet microhabitat use, because they tend to prefer wide-open areas rather than highly vegetated ones in order to forage and feed. Measuring in percentage makes this variable continuous, and this data could be collected via quadrat estimates.

# Problem 2

## a. Clean and wrangle

```{r}
#| label: cleaning-data
# creating data frame called `nests_clean`
nests_clean <- nest_data |> 
  # selecting columns of interest
  select(height_cm, bird_species, ant_species) |> 
  # filtering to only include ant species AB, RRB, and BBR
  filter(ant_species == "AB" |
         ant_species == "RRB" |
         ant_species == "BBR") |> 
  # reordering ant species
  mutate(ant_species = factor(ant_species),
         ant_species = fct_relevel(
         ant_species,
         "AB", "RRB", "BBR")) |> 
  # creating new column to store scientific names
  mutate(scientific_name = case_when(
    ant_species == "AB" ~ "C. sjostedti",
    ant_species == "RRB" ~ "C. mimosae",
    ant_species == "BBR" ~ "C. nigriceps",
    # fills in NA to match with ant species NA
    TRUE ~ NA_character_
  ))
```

## b. Response variable

## c. Purpose of study

## d. Table of models

## e. Run the models

## f. Select the best model

## g. Check the diagnostics

## h. Visualize the model predictions 

## i. Write a caption for your figure

## j. Calculate model predictions

## k. Interpret your results


